---
title: "BIC metabolomics data analysis"
output:
  pdf_document: 
    number_sections: true
  html_notebook: default
---

In this document we present the joint analysis of the different metabolomics datasets submitted to BIC (currently, as of June 2019). We rely on the following resources for metabolomics-specific issues (in addition to the MOP):
Gorrochategui et al. (https://www.sciencedirect.com/science/article/pii/S0165993616300425), section 3.2.5 on data intensity normalization. Li et al. (https://academic.oup.com/nar/article/45/W1/W162/3835313) for description of normalization tools. 

Briefly, metabolic data are senitive to drift and batch effects, especially in long-term studies. To cope with these issues, studies usually use quality control samples throughout the project for assessing the batches, and internal standards and/or quality control metabolites for normalization. Notable methods here are RUV and CCMN. Downstream normalization is used to either deal with the heteroscedasticity of the metabolite-level data (e.g., Pareto normalization) or to account for sample-to-sample variation (MSTUS, Quantile, Median, etc.).

Load the parsed meta-data (from the cloud), required for all analyses presented here.

```{r}
system(paste("~/google-cloud-sdk/bin/gsutil",
             "cp gs://bic_data_analysis/pass1a/pheno_dmaqc/merged_dmaqc_data.RData",
             "."))
load("merged_dmaqc_data.RData")
system("rm merged_dmaqc_data.RData")
rownames(merged_dmaqc_data) = as.character(merged_dmaqc_data$viallabel)

# load required libraries
library(ggplot2);library(reshape2);library(gridExtra)

# load our helper functions
source("~/Desktop/repos/motrpac/tools/preprocessing_helper_functions.R")
source("~/Desktop/repos/motrpac/tools/gcp_functions.R")
```

# Load all datasets

```{r}
metabolomics_data_bucket = "gs://bic_data_analysis/pass1a/metabolomics/"
rdata_files = get_files_in_bucket(metabolomics_data_bucket)
rdata_files = rdata_files[grepl(".rdata$",rdata_files,ignore.case = T)]
site_names = c("duke","mayo","broad","gtech")
for(site in site_names){
  inds = grepl(site,rdata_files,ignore.case = T)
  names(rdata_files)[inds] = site
}
rdata_files = gsub(metabolomics_data_bucket,"",rdata_files)
rdata_objects = sapply(rdata_files,load_from_bucket,
                       bucket=metabolomics_data_bucket,delete=T)
sapply(rdata_objects,names)
newl = list()
for(nn in names(rdata_objects)){
  for(nn2 in names(rdata_objects[[nn]])){
    newl[[nn2]] = rdata_objects[[nn]][[nn2]]
  }
}
rdata_objects = newl

# Separate the datasets by site, tissue, and type (metadata or not)
raw_abundance_datasets =  rdata_objects[grepl("comp",names(rdata_objects))]
names(raw_abundance_datasets) = gsub("_compounds","",names(raw_abundance_datasets))
sample_metadata = rdata_objects[grepl("sampleme",names(rdata_objects))]
names(sample_metadata) = gsub("_samplemeta","",names(sample_metadata))
get_tissue_from_dataset_name<-function(x){return(strsplit(x,split="_")[[1]][2])}
raw_abundance_datasets_tissues = 
  sapply(names(raw_abundance_datasets),get_tissue_from_dataset_name)
get_site_from_dataset_name<-function(x){return(strsplit(x,split="_")[[1]][1])}
raw_abundance_datasets_sites = 
  sapply(names(raw_abundance_datasets),get_site_from_dataset_name)


# Create datasets for the comparison: remove controls, normalize, fold changes
# this is a named list of dataset
# each name has: site,tissue
# each entry has abundance data, row annot data, control data, sample metadata (optional)
metabolomics_parsed_datasets = list()
for(dataset in names(raw_abundance_datasets)){
  site = raw_abundance_datasets_sites[[dataset]]
  tissue_name = raw_abundance_datasets_tissues[[dataset]]
  is_untargeted = is.element(site,set = c("broad","gtech"))
  d = raw_abundance_datasets[[dataset]]
  sample_meta = sample_metadata[[dataset]]
  if(is_untargeted){
    row_annot = d[,1:8];d = d[-c(1:8)]
  }
  else{
    row_annot = d[,1:2];d = d[-c(1:2)]
  }
    
  d[is.na(d)] = 0
  d = log(d+1,base=2)
  d = d[apply(d,1,sd)>0,]
    
  control_columns = !is.element(colnames(d),set=merged_dmaqc_data$viallabel)
  control_data = d[,control_columns]
  d  = d[,!control_columns]
    
  # if site is broad then quantile normalize
  if(is_untargeted){
    d = run_quantile_normalization(d)
  }
    
  print(paste("excluded rows with zero variance",sum(apply(d,1,sd)==0)))
  rowinds = apply(d,1,sd)>0
  d = d[rowinds,]
  row_annot = row_annot[rowinds,]
  control_data=control_data[rowinds,]
    
  # # Compute fold change data vs. the mean pre treatment
  # curr_time_points = merged_dmaqc_data[colnames(d),"animal.key.timepoint"]
  # curr_controls = merged_dmaqc_data[colnames(d),"animal.key.is_control"]
  # raw_group_info = merged_dmaqc_data[colnames(d),"animal.key.anirandgroup"]
  # curr_sex = merged_dmaqc_data[colnames(d),"animal.registration.sex"]
    
  obj_name = paste(site,tissue_name,sep="_")
  obj = list(sample_data=d,control_data=control_data,
             row_annot=row_annot,sample_meta=sample_meta)
  metabolomics_parsed_datasets[[dataset]] = obj
}
```

# Pairwise comparisons

## Using simple correlations

## Using PLS2 regression

```{r,fig.align="center",out.height='60%',out.width='60%',eval=FALSE}
y = metabolomics_parsed_datasets$duke_muscle$sample_data
x = metabolomics_parsed_datasets$broad_muscle$sample_data

y_vials = colnames(y)
bid_y = merged_dmaqc_data[colnames(y),"bid"]
bid_x = merged_dmaqc_data[colnames(x),"bid"]
table(is.element(bid_y,set=bid_x))
colnames(x) = bid_x
colnames(y) = bid_y
x = x[,colnames(y)] # make x in the same order as y
x = t(as.matrix(x))
y = t(as.matrix(y))
y_pca = prcomp(y)
plot(y_pca)
explained_var = y_pca$sdev^2/sum(y_pca$sdev^2)
y_pca_matrix = y_pca$x[,1:10]

# regress out sex, weight
cov_cols = c("animal.registration.sex",
             "acute.test.weight",
             "acute.test.distance")
covs = merged_dmaqc_data[y_vials,cov_cols]

get_explained_variance_using_PCA(x,y)
x = apply(x,2,regress_out,covs=covs)
y = apply(y,2,regress_out,covs=covs)
get_explained_variance_using_PCA(x,y)

library(pls)
library(gplots)
pls_model = plsr(y_pca_matrix~x,ncomp = 10)
r2s = pls::R2(pls_model,ncomp=10)$val[1,,]
plot(explained_var[1:10],r2s[,2])

```

## Using canonical correlations
```{r,fig.align="center",out.height='60%',out.width='60%',eval=FALSE}
cca_res = whitening::scca(x,y,scale=F,lambda.cor = 0.9)
ccaX = tcrossprod( x, cca_res$WX)
ccaY = tcrossprod( y, cca_res$WY)

plot(ccaX[,1],ccaX[,2])
plot(ccaY[,1],ccaY[,2])

```

## Comparison of covariance matrices

```{r}

CV <-function(x){
  return(sd(x)/mean(x))
}

# Get all correlation and covariance matrices
y = metabolomics_parsed_datasets[[1]]$sample_data # anchor all datasets by these ids
y_vials = colnames(y)
bid_y = as.character(merged_dmaqc_data[colnames(y),"bid"])
cov_cols = c("animal.registration.sex",
             "acute.test.weight",
             "acute.test.distance")
covs = merged_dmaqc_data[y_vials,cov_cols]

cov_matrices = list()
cor_matrices = list()
for(nn in names(metabolomics_parsed_datasets)){
  x = metabolomics_parsed_datasets[[nn]]$sample_data
  bid_x = as.character(merged_dmaqc_data[colnames(x),"bid"])
  print(sum(!is.element(bid_y,set=bid_x))) # should be zero
  colnames(x) = bid_x
  x = x[,bid_y]
  x = apply(x,1,regress_out,covs=covs)
  x = t(x)
  cvs = apply(x,1,CV)
  print(table(cvs>0.5))
  x = x[cvs>0.5,]
  # pcax = t(prcomp(t(x),scale. = F)$x[,1:10])
  cov_matrices[[nn]] = cov(x)
  cor_matrices[[nn]] = cor(x)
}
sapply(cor_matrices,dim)
library('evolqg');library(corrplot)
mantel_tests= MantelCor(cor_matrices)
mcors= MatrixCor(cor_matrices)
# diag(mantel_tests)=0
corrplot(mcors,p.mat=mantel_tests$probabilities,type="lower",
         insig = "label_sig",method="shade",
         sig.level = c(.0001, .001, .01),
         pch.cex = .9, pch.col = "black",tl.cex=0.6)
rs_res = RandomSkewers(cor_matrices)
corrplot(rs_res$correlations,
         p.mat=mantel_tests$probabilities,type="lower",
         insig = "label_sig",method="shade",
         sig.level = c(.0001, .001, .01),
         pch.cex = .9, pch.col = "black",tl.cex=0.6)

library(psych)

r1 = cor_matrices[[1]]
r2 = cor_matrices[[8]]
cor(r1[lower.tri(r1)],r2[lower.tri(r2)])
mantel_tests[1,8]
mantel_tests[8,1]

threshold_comp = apply_function_on_pairs(cor_matrices,
          function(x,y)sum(x>0.7&y>0.7))
corrplot(threshold_comp,is.corr = F,order="hclust")

threshold_comp = apply_function_on_pairs(cor_matrices,
         function(x,y)mean(diag(cor(x,y))))
corrplot(threshold_comp,is.corr = F,order="hclust")

```

# Integrarion with RNAseq

# Differential abundance analysis












