---
title: "BIC metabolomics data analysis"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document:
    number_sections: yes
---

In this document we present the joint analysis of the PASS1A metabolomics datasets.

# Load all datasets

Load the data from the cloud, including: phenotypic data, metabolomic datasets, and metabolomics dictionary.

```{r,results='hide',message=FALSE,warning=FALSE}
# load the data and functions to the session

# helper functions for unsupervised data matrix normalization
source("~/Desktop/repos/motrpac-bic-norm-qc/tools/unsupervised_normalization_functions.R")
# basic functions for reading data objects from GCP
source("~/Desktop/repos/motrpac-bic-norm-qc/tools/gcp_functions.R")
# helper functions for data reformat/reshape
source("~/Desktop/repos/motrpac-bic-norm-qc/tools/data_aux_functions.R")
# helper functions for reading and checking the metabolomics data from the cloud
source("~/Desktop/repos/motrpac-bic-norm-qc/tools/metabolomics_data_parsing_functions.R")
# functions for association analysis
source("~/Desktop/repos/motrpac-bic-norm-qc/tools/association_analysis_methods.R")
# functions for classification/regression and cross-validation
source("~/Desktop/repos/motrpac/tools/prediction_ml_tools.R")

library(randomForest) # for classification tests
library(VennDiagram) # for overlaps
require(gridExtra)
library(corrplot)

# Load the dmaqc data
merged_dmaqc_data =  load_from_bucket("merged_dmaqc_data2019-10-15.RData",
    "gs://bic_data_analysis/pass1a/pheno_dmaqc/",T)
merged_dmaqc_data = merged_dmaqc_data[[1]]
rownames(merged_dmaqc_data) = as.character(merged_dmaqc_data$vial_label)
# define the tissue variable
merged_dmaqc_data$tissue = merged_dmaqc_data$sampletypedescription
# define the time to freeze variable
merged_dmaqc_data$time_to_freeze = merged_dmaqc_data$calculated.variables.time_death_to_collect_min + 
  merged_dmaqc_data$calculated.variables.time_collect_to_freeze_min

# blood freeze times
blood_samples = 
  merged_dmaqc_data$specimen.processing.sampletypedescription ==
  "EDTA Plasma"
blood_freeze_time = 
  as.difftime(merged_dmaqc_data$specimen.processing.t_freeze,units = "mins") -
  as.difftime(merged_dmaqc_data$specimen.processing.t_edtaspin,units="mins")
blood_freeze_time = as.numeric(blood_freeze_time)
time_to_freeze = merged_dmaqc_data$time_to_freeze[blood_samples] = 
  blood_freeze_time[blood_samples]

# Load our parsed metabolomics datasets
metabolomics_bucket_obj = load_from_bucket(
  file = "metabolomics_parsed_datasets_pass1a_external_v2_04112020.RData",
  bucket = "gs://bic_data_analysis/pass1a/metabolomics/")
metabolomics_parsed_datasets = metabolomics_bucket_obj$metabolomics_parsed_datasets

# destination for selected figures, tables and RData files
local_destination = "~/Desktop/pass1a_met_files/" # change if you want another dir
try(dir.create(local_destination))
analysis_out_files_dest = paste(local_destination,"/analysis_out_files/",sep="")
try(dir.create(analysis_out_files_dest))

```

Define the parameters for the analyses below including which covariates to take and what normalization methods to consider.

```{r}
# Define the variables to be adjusted for:
biospec_cols = c(
  "acute.test.distance",
  "calculated.variables.time_to_freeze",
  "bid" # required for matching datasets
  )
differential_analysis_cols = c(
  "animal.registration.sex",
  "animal.key.timepoint",
  "animal.key.is_control",
  "sample_order"
)
# normalization method pairs to test
# first is for targeted datsets, second is for untargeted
tar_vs_untar_norm_pairs = list(
  c("log2,imp","log,med,imp"),
  c("imp,none","imp,none"),
  c("imp,none","log,med,imp")
)
# specify time points for pairwise differential analysis
tp_pairs_for_de = data.frame(
  tp1 = c(0.0,1.0,4.0,4.0,24.0,7.0),
  tp1_is_control = c(F,F,F,F,F,F),
  tp2 = c(0.0,0.0,1.0,7.0,0.0,7.0),
  tp2_is_control = c(T,F,F,T,T,T)
)
```

# Preprocess the data to ease the analyses below

```{r}
# load the data
metabolomics_processed_datasets = load_from_bucket(
  file="metabolomics_processed_datasets04072020.RData",
  bucket = "gs://bic_data_analysis/pass1a/metabolomics/"
)[[1]]

#' Transform the row names to motrpac names whenever possible.
get_motrpac_dict_rownames<-function(x,row_annot_x){
  if(length(setdiff(rownames(x),row_annot_x[,1]))>0){
    stop("rownames in data and annotation do not match")
  }
  if(length(setdiff(row_annot_x[,1],rownames(x)))>0){
    warning("annotation has additional metabolites")
  }
  if(! "motrpac_comp_name" %in% colnames(row_annot_x)){
    stop("no motrpac_comp_name column in annotation")
  }
  row_annot_x = row_annot_x[rownames(x),]
  y = row_annot_x[,"motrpac_comp_name"]
  inds = !is.na(y) & y!=""
  rnames = rownames(x)
  rnames[inds] = y[inds]
  return(rnames)
}

# Reduce the metadata to the selected columns, transform row names 
# to motrpac ids and add bid-based names
for(currname in names(metabolomics_processed_datasets)){
  d = metabolomics_processed_datasets[[currname]]
  curr_data = d$normalized_data[[1]]
  # organize the metadata
  # some covariates may not appear in the data because they had
  # too many NAs
  covered_covs = intersect(union(biospec_cols,differential_analysis_cols),
          colnames(d$sample_meta))
  curr_meta = d$sample_meta[colnames(curr_data),covered_covs]
  # remove metadata variables with too many NAs
  na_counts = apply(is.na(curr_meta),2,sum)
  curr_meta = curr_meta[,na_counts/nrow(curr_meta) < 0.1]
  motrpac_met_names = get_motrpac_dict_rownames(d$normalized_data[[1]],d$row_annot)
  names(motrpac_met_names) = rownames(d$normalized_data[[1]])
  # add the new fields
  d$covs = curr_meta
  d$motrpac_met_names = motrpac_met_names
  metabolomics_processed_datasets[[currname]] = d
}

# Transform sample id to bid
for(currname in names(metabolomics_processed_datasets)){
  d = metabolomics_processed_datasets[[currname]]
  bids = as.character(d$sample_meta[colnames(d$normalized_data[[1]]),"bid"])
  if(any(bids!=d$sample_meta[,"bid"])){
    print(paste("Warning in dataset",
           currname,"sample order in data and metadata do not match"))
  }
  if(length(unique(bids))==ncol(d$normalized_data[[1]])){
    rownames(d$sample_meta) = bids
    for(j in 1:length(d$normalized_data)){
      colnames(d$normalized_data[[j]]) = bids
    }
  }
  else{
    print(paste("there are ",nrow(d$sample_meta),"samples in dataset",currname))
    print(paste(" => bid column names are not unique in:",currname,", merging"))
    oldmeta = d$sample_meta
    newmeta_inds = c()
    for(bid in unique(bids)){
      newmeta_inds = c(newmeta_inds,which(bids==bid)[1])
    }
    newmeta = oldmeta[newmeta_inds,]
    rownames(newmeta) = as.character(newmeta[,"bid"])
    d$sample_meta = newmeta
    for(j in 1:length(d$normalized_data)){
      d$normalized_data[[j]] = aggregate_repeated_samples(
        d$normalized_data[[j]],bids,na.rm=T
      )
      if(!(all(colnames(d$normalized_data[[j]]) %in%
               rownames(newmeta)))){
        print(paste("!!! Error in merging samples by bid in dataset",currname))
      }
      d$normalized_data[[j]] = d$normalized_data[[j]][
        ,rownames(newmeta)]
    }
    d$motrpac_met_names = d$motrpac_met_names[rownames(d$normalized_data[[1]])]
    print("done")
  }
  metabolomics_processed_datasets[[currname]] = d
  
}

# Get the named datasets separately - easier to work with later
processed_named_datasets = list()
for(currname in names(metabolomics_processed_datasets)){
  d = metabolomics_processed_datasets[[currname]]
  inds = !is.na(d$row_annot$motrpac_comp_name)
  d$row_annot = d$row_annot[inds,]
  d$motrpac_met_names = d$motrpac_met_names[inds]
  if(!all(d$motrpac_met_names == d$row_annot$motrpac_comp_name)){
      print(paste("Error: row name problems in dataset ",currname))
  }
  for(j in 1:length(d$normalized_data)){
    d$normalized_data[[j]] = d$normalized_data[[j]][inds,]
    if(!all(rownames(d$normalized_data[[j]]) == rownames(d$row_annot))){
      print(paste("Error: row name problems in dataset ",currname))
    }
    rownames(d$normalized_data[[j]]) = d$motrpac_met_names
  }
  processed_named_datasets[[currname]] = d
}
gc()
```


# Compound overlap among the platforms

Compare the overlap between the named metabolites. This is computed for all dataset pairs, but tissue-specific plots are presented for simplicity.

```{r}
dataset2named_metabolites = lapply(processed_named_datasets,
       function(x)rownames(x$normalized_data[[1]]))
Nd = length(processed_named_datasets)
overlap_matrix = matrix(0,Nd,Nd,dimnames = list(
  names(processed_named_datasets),names(processed_named_datasets)
))
pairwise_shared_metabolites = list()
for(nn1 in names(processed_named_datasets)){
  pairwise_shared_metabolites[[nn1]] = list()
  for(nn2 in names(processed_named_datasets)){
    shared_metabolites = intersect(
      dataset2named_metabolites[[nn1]],dataset2named_metabolites[[nn2]])
    pairwise_shared_metabolites[[nn1]][[nn2]] = shared_metabolites
    overlap_matrix[nn1,nn2] = length(shared_metabolites)
  }
}

dataset2is_targeted = 
  sapply(processed_named_datasets,function(x)x$is_targeted)
dataset2tissue = 
  sapply(processed_named_datasets,function(x)x$tissue)
tissue2overlap_matrix = list()
met_lists_coverage = c()
for(tissue in unique(dataset2tissue)){
  currdatasets = names(dataset2tissue)[dataset2tissue==tissue]
  if(length(currdatasets)<2){next}
  tissue2overlap_matrix[[tissue]] = 
    overlap_matrix[currdatasets,currdatasets]
  all_tissue_mets = unique(unlist(dataset2named_metabolites))
  all_targeted_mets = unique(
    unlist(dataset2named_metabolites[dataset2is_targeted & dataset2tissue==tissue])
  )
  all_untargeted_mets = unique(
    unlist(dataset2named_metabolites[!dataset2is_targeted & dataset2tissue==tissue])
  )
  currv = rep("tar_only",length(all_tissue_mets))
  in_tar = all_tissue_mets %in% all_targeted_mets
  in_untar = all_tissue_mets %in% all_untargeted_mets
  currv[in_untar] = "untar_only"
  currv[in_untar & in_tar] = "both"
  currv = paste(tissue,currv,sep=",")
  met_lists_coverage = rbind(met_lists_coverage,
                             cbind(all_tissue_mets,currv))
}

save(overlap_matrix,tissue2overlap_matrix,met_lists_coverage,
     dataset2named_metabolites, pairwise_shared_metabolites,
    file = paste(analysis_out_files_dest,
                 "plarform_named_metabolite_overlaps.RData",sep=""))

# Show the pairwise overlap and the overall Venn diagram
# comparing targeted to untargeted
for (tissue in names(tissue2overlap_matrix)){
  rownames(tissue2overlap_matrix[[tissue]]) = gsub(
    paste0(tissue,","),"",rownames(tissue2overlap_matrix[[tissue]])
  )
  colnames(tissue2overlap_matrix[[tissue]]) = rownames(tissue2overlap_matrix[[tissue]])
  corrplot(tissue2overlap_matrix[[tissue]],
           is.corr = F,method = "number",number.cex = 0.7,cl.length = 10)
  m = met_lists_coverage[grepl(tissue,met_lists_coverage[,2]),]
  inter_area = sum(grepl("both",m[,2]))
  tar_area = sum(grepl(",tar_only",m[,2])) + inter_area
  untar_area = sum(grepl(",untar_only",m[,2])) + inter_area
  venng = draw.pairwise.venn(tar_area,untar_area,inter_area,
            c("Targeted","Untargeted"),lty = rep("blank",2), 
            fill = c("pink", "cyan"), alpha = rep(0.5, 2),
            cat.dist = rep(0.01, 2),ind=F)
  grid.arrange(gTree(children=venng),
               top=paste(tissue,"measured named metabolites",sep=", "))
}

```

# Correlations of shared metabolites

```{r,out.height='50%',out.width='50%',message=FALSE,warning=FALSE}
# For each normalization option (a pair from the list above) we compute
# all pairwise correlation matrices of the overlapping metabolites
norm_method2correlations = list()
for(norm_pairs in tar_vs_untar_norm_pairs){
  tar_norm_method = norm_pairs[1]
  untar_norm_method = norm_pairs[2]
  metabolite_corrs = list()
  for(nn1 in names(processed_named_datasets)){
    nn1_tissue = processed_named_datasets[[nn1]]$tissue
    if(grepl(",tar,",nn1)){nn1_norm  = tar_norm_method}
    if(grepl(",untar,",nn1)){nn1_norm  = untar_norm_method}
    metabolite_corrs[[nn1]] = list()
    for(nn2 in names(processed_named_datasets)){
      if(nn2 == nn1){next}
      if(grepl(",tar,",nn2)){nn2_norm  = tar_norm_method}
      if(grepl(",untar,",nn2)){nn2_norm  = untar_norm_method}
      nn2_tissue = processed_named_datasets[[nn2]]$tissue
      nn2_dataset = paste(strsplit(nn2,split=",")[[1]][3:4],collapse = ",")
      if(nn1_tissue!=nn2_tissue){next}
      # get the numeric datasets and their annotation
      x = processed_named_datasets[[nn1]]$normalized_data[[nn1_norm]]
      y = processed_named_datasets[[nn2]]$normalized_data[[nn2_norm]]
      x = x[apply(x,1,sd)>0,]
      y = y[apply(y,1,sd)>0,]
      # compute correlations on overlapping bids
      currbids = intersect(colnames(x),colnames(y))
      x = x[,currbids];y=y[,currbids]
      
      # Transform to original scale if one dataset is not log-transformed
      is_tar_log = grepl("log",tar_norm_method)
      is_untar_log = grepl("log",untar_norm_method)
      if(is_tar_log && !is_untar_log){x = 2^x}
      if(!is_tar_log && is_untar_log){y = 2^y}
    
      # If data are not log-transformed then scale the rows
      if(!is_tar_log || !is_untar_log){
        x = t(scale(t(x),center = T,scale = T))
        y = t(scale(t(y),center = T,scale = T))
      }
    
      # Compute the correlation matrices of the shared metabolites
      shared_metabolites = intersect(rownames(x),rownames(y))
      if(length(shared_metabolites)==0){next}
      if(length(shared_metabolites)>1){
          metabolite_corrs[[nn1]][[nn2]] = cor(t(x[shared_metabolites,]),
                t(y[shared_metabolites,]),method = "spearman")
      }
      else{
          metabolite_corrs[[nn1]][[nn2]] = cor(x[shared_metabolites,],
                y[shared_metabolites,],method = "spearman")
      }
      if(any(is.na(metabolite_corrs[[nn1]][[nn2]]))){
        print("NAs in correlation computation")
        print(paste(nn1,"vs",nn2))
        print(table(sum(x)))
        print(table(sum(y)))
      }
    }
  }
  pair_name = paste(tar_norm_method,untar_norm_method,sep=";")
  norm_method2correlations[[pair_name]] = metabolite_corrs
}

# Plot some correlations
rep_correlations = c()
tissues = unique(sapply(processed_named_datasets,function(x)x$tissue))
for(pair_name in names(norm_method2correlations)){
  metabolite_corrs =  norm_method2correlations[[pair_name]]
  for(tissue in tissues){
    curr_datasets = names(metabolite_corrs)[
      grepl(tissue,names(metabolite_corrs))
    ]
    between_corrs = c()
    for(dataset in curr_datasets){
      l = metabolite_corrs[[dataset]]
      between_corrs = c(between_corrs,unname(unlist(sapply(l,diag))))
    }
    rep_correlations = rbind(rep_correlations,
          c(pair_name,tissue,mean(between_corrs,na.rm=T),sd(between_corrs,na.rm=T)))
  }
}
rep_correlations = rep_correlations[!is.na(rep_correlations[,3]),]
rep_correlations = data.frame(
  "NormMethod" = rep_correlations[,1],
  "Tissue" = rep_correlations[,2],
  "Mean" = as.numeric(rep_correlations[,3]),
  "SD" = as.numeric(rep_correlations[,4])
)
rep_correlations = rep_correlations[order(rep_correlations$Tissue),]
print(
    ggplot(rep_correlations, aes(x=Tissue, y=Mean, fill=NormMethod)) +
      geom_bar(position=position_dodge(), stat="identity", colour='black') +
      geom_errorbar(aes(ymin=Mean-SD, ymax=Mean+SD),na.rm=T, 
                   width=.2,position=position_dodge(.9))
)
```

We examine the average correlation between the platforms (within tissues). Whenever two platforms share more than a single metabolite we plot both the average correlation between the same metabolites and between other metabolites. Adding the average correlation between platforms but with different metabolites is important as it gives some perspective to what a significant correlation is. That is, in many cases below, the average correlation may be greater than expected.

```{r,out.height='50%',out.width='50%',message=FALSE,warning=FALSE}
# Next examine the Spearman correlations between platforms
mean_abs<-function(x,...){return(mean(abs(x),...))}
sd_abs<-function(x,...){return(sd(abs(x),...))}
extract_diag_vs_non_diag<-function(corrs,func=mean,...){
  if(length(corrs)==1){
    return(c(same=func(corrs,...),other=NA))
  }
  same = func(diag(corrs),...)
  other = func(
    c(corrs[lower.tri(corrs,diag = F)]),...)
  return(c(same=same,other=other))
}

metabolite_corrs = norm_method2correlations$`imp,none;imp,none`
for(tar_dataset in names(metabolite_corrs)){
  if(!grepl(",tar,",tar_dataset)){next}
  l = metabolite_corrs[[tar_dataset]]
  if(length(l)==0){next}
  corr_info = as.data.frame(t(sapply(l, extract_diag_vs_non_diag)))
  corr_sd = as.data.frame(t(sapply(l, extract_diag_vs_non_diag,func=sd)))
  # shorten the row names
  rownames(corr_info) = sapply(rownames(corr_info),
      function(x)paste(strsplit(x,split=",")[[1]][3:4],collapse=","))
  rownames(corr_sd) = rownames(corr_info)
  corr_info$dataset = rownames(corr_info)
  corr_sd$dataset = corr_info$dataset
  corr_info = melt(corr_info)
  corr_sd = melt(corr_sd)
  corr_info$sd = corr_sd$value
  print(
    ggplot(corr_info, aes(x=dataset, y=value, fill=variable)) +
      geom_bar(position=position_dodge(), stat="identity", colour='black') +
      geom_errorbar(aes(ymin=value-sd, ymax=value+sd),na.rm=T, 
                   width=.2,position=position_dodge(.9)) +
    ggtitle(tar_dataset) + xlab("Untargeted dataset") + ylab("Spearman") +
      labs(fill = "Pair type") + 
      theme(legend.position="top",legend.direction = "horizontal")
  )
}
```

# Differential analysis

```{r,out.height='50%',out.width='50%',message=FALSE,warning=FALSE}
# For each normalization option (a pair from the list above) we compute
# all pairwise correlation matrices of the overlapping metabolites
single_metabolite_de = c()
for(currname in names(processed_named_datasets)){
  d = processed_named_datasets[[currname]]
  for(norm_method in names(d$normalized_data)){
    x = d$normalized_data[[norm_method]]
    # if data was not log-transformed then scale it
    if(!grepl("log",norm_method)){
      x = t(scale(t(x),center = T,scale = T))
    }
    x_meta = d$sample_meta
    covs = data.frame(
      sex = x_meta$animal.registration.sex,
      sample_order = x_meta$sample_order
    )
    for(j in 1:nrow(tp_pairs_for_de)){
      currres = t(apply(
        x,1,simple_pairwise_differential_abundance,
        tps = x_meta$animal.key.timepoint,
        is_control = x_meta$animal.key.is_control,
        tp1 = tp_pairs_for_de[j,1],tp1_iscontrol = tp_pairs_for_de[j,2],
        tp2 = tp_pairs_for_de[j,3],tp2_iscontrol = tp_pairs_for_de[j,4],
        covs = covs
      ))
      currres  = data.frame(currres)
      currres$dataset = currname
      currres$is_targeted = d$is_targeted
      currres$tissue = d$tissue
      currres$norm = norm_method
      currres$is_log_scale = grepl(norm_method,"log")
      currres$metabolite = rownames(currres)
      rownames(currres) = NULL
      single_metabolite_de = rbind(single_metabolite_de,currres)
    }
    # repeat the calculation if data was logged
    if(grepl(norm_method,"log")){
      x = 2^x
      x = t(scale(t(x),center = T,scale = T))
      for(j in 1:nrow(tp_pairs_for_de)){
        currres = t(apply(
          x,1,simple_pairwise_differential_abundance,
          tps = x_meta$animal.key.timepoint,
          is_control = x_meta$animal.key.is_control,
          tp1 = tp_pairs_for_de[j,1],tp1_iscontrol = tp_pairs_for_de[j,2],
          tp2 = tp_pairs_for_de[j,3],tp2_iscontrol = tp_pairs_for_de[j,4],
          covs = covs
        ))
        currres  = data.frame(currres)
        currres$dataset = currname
        currres$is_targeted = d$is_targeted
        currres$tissue = d$tissue
        currres$norm = norm_method
        currres$is_log_scale = F
        currres$metabolite = rownames(currres)
        rownames(currres) = NULL
        single_metabolite_de = rbind(single_metabolite_de,currres)
      }
    }
  }
}
```



```{r,out.height='50%',out.width='50%',message=FALSE,warning=FALSE}

library(metafor)
for(pair_name in names(norm_method2comparison_results)){
  meta_analysis_stats = list()
  single_metabolite_de = 
    norm_method2comparison_results[[pair_name]]$single_metabolite_de
  for(tissue in unique(single_metabolite_de$tissue)){
    for(tp in unique(single_metabolite_de$tp)){
      curr_subset = single_metabolite_de[
        single_metabolite_de$tissue==tissue &
          single_metabolite_de$tp==tp,]
      for(metabolite in unique(curr_subset$metabolite)){
        curr_met_data = curr_subset[
          curr_subset$metabolite==metabolite,]
        curr_met_data$var = curr_met_data$Std^2
        re_model1 = NULL;re_model2=NULL
        re_model_tar = NULL;re_model_untar = NULL
        try({re_model1 = rma.uni(curr_met_data$Est,curr_met_data$var,method="FE")})
        try({re_model2 = rma.mv(curr_met_data$Est,curr_met_data$var,
                           mods=curr_met_data$is_targeted,method="FE")})
        try({re_model_tar = rma.uni(
          curr_met_data[curr_met_data$is_targeted==1,"Est"],
          curr_met_data[curr_met_data$is_targeted==1,"var"],
          method="FE"
        )})
        try({re_model_untar = rma.uni(
          curr_met_data[curr_met_data$is_targeted==0,"Est"],
          curr_met_data[curr_met_data$is_targeted==0,"var"],
          method="FE"
        )})
        meta_analysis_stats[[paste(metabolite,tissue,tp,sep=";")]] = 
          list(curr_met_data=curr_met_data,re_model1=re_model1,
            re_model2 = re_model2,re_model_tar=re_model_tar,
            re_model_untar = re_model_untar)
      }
    }
  }
  norm_method2comparison_results[[pair_name]]$meta_analysis_stats =
    meta_analysis_stats
}

save(norm_method2comparison_results,
    file = paste(analysis_out_files_dest,
                 "norm_method2comparison_results.RData",sep=""))

```


```{r,out.height='50%',out.width='50%',message=FALSE,warning=FALSE}
pthr = 0.001
for(tissue in tissues){
  for(pair_name in names(norm_method2comparison_results)){
      meta_analysis_stats = 
          norm_method2comparison_results[[pair_name]]$meta_analysis_stats
      meta_analysis_stats = meta_analysis_stats[
        grepl(tissue,names(meta_analysis_stats))
      ]
      if(length(meta_analysis_stats)==0){next}
      naive_analysis_tar = sapply(meta_analysis_stats,
        function(x)sum(x$curr_met_data[,"Pvalue"]<pthr &
                     x$curr_met_data[,"is_targeted"]))
      naive_analysis_untar = sapply(meta_analysis_stats,
        function(x)sum(x$curr_met_data[,"Pvalue"]<pthr &
                     !x$curr_met_data[,"is_targeted"]))
      tb = table(naive_analysis_tar,naive_analysis_untar)
      nonsig_in_both = tb[1,1]
      tb[1,1] = 0
      barplot(t(tb),
              legend=T,xlab="Number of targeted datasets with p<0.001",
              ylab = "number of metabolites",
              main = paste(tissue,pair_name))
      
      # write.table(
      #   names(which(naive_analysis_tar>0 & naive_analysis_untar==0)),
      #   quote = F,row.names = F
      # )
      # write.table(
      #   names(which(naive_analysis_tar==0 & naive_analysis_untar>0)),
      #   quote = F,row.names = F
      # )
  }
}
```

Meta-analysis results

```{r,out.height='50%',out.width='50%',message=FALSE,warning=FALSE}
pthr = 0.001
tissues = unique(sapply(metabolomics_processed_datasets,function(x)x$tissue))
tar_vs_untar_correlations = c()
for(tissue in tissues){
  for(pair_name in names(norm_method2comparison_results)){
    meta_analysis_stats = 
        norm_method2comparison_results[[pair_name]]$meta_analysis_stats
    meta_analysis_stats  = meta_analysis_stats[
        grepl(tissue,names(meta_analysis_stats))
      ]
    if(is.null(meta_analysis_stats) || length(meta_analysis_stats)==0){next}
  
    # P-value for the difference between targeted and untargeted
    targeted_diff_p = 
    sapply(meta_analysis_stats,function(x)x$re_model2$pval[2])

   # P-values - targeted vs. untargeted
    pvals_tar = sapply(meta_analysis_stats,function(x)x$re_model_tar$pval)
    pvals_untar = sapply(meta_analysis_stats,function(x)x$re_model_untar$pval)
    pvals_untar = unlist(pvals_untar[sapply(pvals_untar,length)>0])
    significant_in = rep("None",length(pvals_untar))
    significant_in[pvals_tar<pthr] = "Targeted"
    significant_in[pvals_untar<pthr] = "Untargeted"
    significant_in[pvals_tar<pthr & pvals_untar<pthr] = "Both"
    significant_diff = targeted_diff_p<pthr
    rho = cor(-log10(pvals_tar),-log10(pvals_untar),method = "pearson")
    # Betas - targeted vs. untargeted
    betas_tar = 
      sapply(meta_analysis_stats,function(x)x$re_model_tar$beta[1,1])
    betas_untar = 
      sapply(meta_analysis_stats,function(x)x$re_model_untar$beta[1,1])
    betas_untar = unlist(betas_untar[sapply(betas_untar,length)>0])
    df = data.frame(
      targeted = betas_tar,
      untargeted = betas_untar,
      significant_in = significant_in,
      significant_diff = significant_diff
    )
    rho_beta = cor(betas_untar,betas_tar,method = "pearson")
    rhop = cor.test(betas_untar,betas_tar,method = "pearson")$p.value
    print(
      ggplot(df, aes(x=targeted, y=untargeted,
                 shape=significant_diff, color=significant_in)) +
      geom_point() + geom_abline(slope=1,intercept = 0) + 
      ggtitle(paste(tissue, pair_name,
                    "effects, rho=:",format(rho_beta,digits=2)))
    )
  
    tar_vs_untar_correlations = rbind(tar_vs_untar_correlations,
                                      c(tissue,pair_name,rho,rho_beta))
   # draw a venn diagram
    inter_area = sum(significant_in=="Both")
    tar_area = sum(significant_in=="Targeted") + inter_area
    untar_area = sum(significant_in=="Untargeted") + inter_area
    subt = paste("Not significant in both:",table(significant_in)["None"])
    
    ss = paste(sample(names(pvals_untar)[significant_in=="Targeted"])[1:10],collapse=";")
    ss = gsub(paste(",",tissue,sep=""),"",ss)
    # ss
    ss = paste(sample(names(pvals_untar)[significant_in=="Untargeted"])[1:10],collapse=";")
    ss = gsub(paste(",",tissue,sep=""),"",ss)
    # ss
    
    venng = draw.pairwise.venn(tar_area,untar_area,inter_area,
            c("Targeted","Untargeted"),lty = rep("blank",2), 
            fill = c("pink", "cyan"), alpha = rep(0.5, 2),
            cat.dist = rep(0.01, 2),ind=F)
    # grid.newpage()
    grid.arrange(gTree(children=venng), 
                 top=paste(tissue,pair_name), bottom=subt)
  }
}
```


## Plot selected examples

Here are the results for lactate in plasma.

Example from a log2 dataset:

```{r,out.height='50%',out.width='50%',message=FALSE,warning=FALSE}
meta_analysis_stats = 
  norm_method2comparison_results$`log2,imp;med,log,imp`$meta_analysis_stats
lact_res = meta_analysis_stats[
  grepl("lact",names(meta_analysis_stats),ignore.case = T) &
    grepl("plasma",names(meta_analysis_stats),ignore.case = T)
]
lact_res_hours = sapply(names(lact_res),
            function(x)as.numeric(strsplit(x,split=";")[[1]][3]))
lact_res = lact_res[order(lact_res_hours)]
for(lact_example in names(lact_res)[1:6]){
  curr_labels = gsub("plasma,","",
                     meta_analysis_stats[[lact_example]][[1]][,1])
  forest(meta_analysis_stats[[lact_example]]$re_model1,
       slab = curr_labels,
       main = lact_example,xlab = "Log fc",
       col = "blue",cex = 1.1)
}

curr_labels = gsub("plasma,","",
                     meta_analysis_stats[["glutamate;plasma;0"]][[1]][,1])
curr_labels = curr_labels[
  meta_analysis_stats[["glutamate;plasma;0"]][[1]]$is_targeted==0]
forest(meta_analysis_stats[["glutamate;plasma;0"]]$re_model_untar,
        slab = curr_labels,
       main = "glutamate,plasma,0",xlab = "Log fc",
       col = "blue",cex = 1.1)

```

Example from a scaled dataset:

```{r,out.height='50%',out.width='50%',message=FALSE,warning=FALSE}
meta_analysis_stats = 
  norm_method2comparison_results$`imp,none;imp,none`$meta_analysis_stats
lact_res = meta_analysis_stats[
  grepl("lact",names(meta_analysis_stats),ignore.case = T) &
    grepl("plasma",names(meta_analysis_stats),ignore.case = T)
]
lact_res_hours = sapply(names(lact_res),
            function(x)as.numeric(strsplit(x,split=";")[[1]][3]))
lact_res = lact_res[order(lact_res_hours)]
for(lact_example in names(lact_res)[1:6]){
  curr_labels = gsub("plasma,","",
                     meta_analysis_stats[[lact_example]][[1]][,1])
  forest(meta_analysis_stats[[lact_example]]$re_model1,
       slab = curr_labels,
       main = lact_example,xlab = "Log fc",
       col = "blue",cex = 1.1)
}

```

We can now check the same analysis for liver:

```{r,out.height='50%',out.width='50%',message=FALSE,warning=FALSE}
lact_res = meta_analysis_stats[
  grepl("lact",names(meta_analysis_stats),ignore.case = T) &
    grepl("liver",names(meta_analysis_stats),ignore.case = T)
]
lact_res_hours = sapply(names(lact_res),
            function(x)as.numeric(strsplit(x,split=";")[[1]][3]))
lact_res = lact_res[order(lact_res_hours)]
for(lact_example in names(lact_res)[1:6]){
  curr_labels = gsub("liver_powder,","",
                     meta_analysis_stats[[lact_example]][[1]][,1])
  forest(meta_analysis_stats[[lact_example]]$re_model1,
       slab = curr_labels,
       main = lact_example,xlab = "Log fc",
       col = "blue",cex = 1.1)
}

```


<!-- We create report by tissue with all forest plots and analysis results. -->

<!-- ```{r} -->
<!-- meta_analysis_tissues = sapply(names(meta_analysis_stats), -->
<!--                                function(x)strsplit(x,split=",")[[1]][2]) -->
<!-- meta_analysis_metabolites = sapply(names(meta_analysis_stats), -->
<!--                                function(x)strsplit(x,split=",")[[1]][1]) -->

<!-- for(tissue in unique(meta_analysis_tissues)){ -->
<!--   pdf(paste("~/Desktop/",tissue,".pdf",sep="")) -->
<!-- } -->

<!-- ``` -->

From the plots above we take the most extreme examples and examine their forest plots.

```{r,out.height='50%',out.width='50%',message=FALSE,warning=FALSE}
meta_analysis_stats = 
  norm_method2comparison_results$`imp,none;imp,none`$meta_analysis_stats

# P-value for the difference between targeted and untargeted
targeted_diff_p = 
  sapply(meta_analysis_stats,function(x)x$re_model2$pval[2])

# P-values - targeted vs. untargeted
pvals_tar = sapply(meta_analysis_stats,function(x)x$re_model_tar$pval)
pvals_untar = sapply(meta_analysis_stats,function(x)x$re_model_untar$pval)
pvals_untar = unlist(pvals_untar[sapply(pvals_untar,length)>0])


agree_example = names(sample(which(pvals_tar< 1e-5 & pvals_untar < 1e-5 &
                                     targeted_diff_p > 0.1))[1])
simplify_labels_for_forest<-function(s){
  s = gsub(",untargeted","",s)
  tissue = strsplit(s,split=",")[[1]][1]
  s = gsub(paste(tissue,",",sep=""),"",s)
  return(s)
}
forest(meta_analysis_stats[[agree_example]]$re_model1,
  slab = simplify_labels_for_forest(
    meta_analysis_stats[[agree_example]][[1]][,1]),
  main = paste(agree_example,"significant in both, tar and untar agree",sep="\n"),
  xlab = "Log fc",col = "blue")

agree_p_disagree_beta = names(sample(which(pvals_tar< 1e-5 & pvals_untar < 1e-5 &
                                     targeted_diff_p < 0.01))[1])
forest(meta_analysis_stats[[agree_p_disagree_beta]]$re_model1,
  slab = simplify_labels_for_forest(
    meta_analysis_stats[[agree_p_disagree_beta]][[1]][,1]),
  main = paste(agree_p_disagree_beta,
               "significant in both, tar and untar disagree",sep="\n"),
  xlab = "Log fc",col = "blue")

disagree_example1 = names(sample(which(pvals_tar< 1e-5 & pvals_untar >0.1))[1])
forest(meta_analysis_stats[[disagree_example1]]$re_model1,
  slab = simplify_labels_for_forest(
    meta_analysis_stats[[disagree_example1]][[1]][,1]),
  main = paste(disagree_example1,
               "significant targeted, tar and untar disagree",sep="\n"),
  xlab = "Log fc",col = "blue")


disagree_example2 = names(sample(which(pvals_tar > 0.1 & pvals_untar < 1e-20))[1])
forest(meta_analysis_stats[[disagree_example2]]$re_model1,
  slab = simplify_labels_for_forest(
    meta_analysis_stats[[disagree_example2]][[1]][,1]),
  main = paste(disagree_example2,
               "significant in untargeted, tar and untar disagree",sep="\n"),
  xlab = "Log fc",col = "blue")

```


<!-- # Log-transform: effect on differential analysis in targeted data -->

<!-- Untargeted data are typically log-transformed and analyzed using linear models. On the other hand, concentration data are sometimes analyzed with the same type of models but using the original data. This raises a problem if we wish to compare exact statistics from these data. In this section we perform residual analysis for single metabolites. Our goal is to identify if concentration data behaves "normally" when not log-transformed. The analysis below examines the residuals of the data after fitting linear models for each metabolite, adjusting for freeze time and sex. We then compare the results with and without the log-transformation, counting the number of metabolites with a significant evidence for non-normally distributed residuals.  -->

<!-- ```{r,out.height='50%',out.width='50%',message=FALSE,warning=FALSE} -->
<!-- # check for normality using the Kolmogorov-Smirnov test -->
<!-- is_normal_test<-function(v){ -->
<!--   if(sd(v,na.rm = T)==0){return(0)} -->
<!--   try({return(shapiro.test(v)$p.value)}) -->
<!--   # The Shapiro test may fail if the sd of v is zero -->
<!--   return(ks.test(v,"pnorm",mean(v,na.rm=T),sd(v,na.rm = T))$p.value) -->
<!-- } -->
<!-- # go over the named datasets, get a logged and an unlogged version of -->
<!-- # the data, use these as inputs for the regression -->
<!-- residual_analysis_results = list() -->
<!-- for(nn1 in names(metabolomics_processed_datasets)){ -->
<!--   if(!metabolomics_processed_datasets[[nn1]]$is_targeted){next} -->
<!--   d = metabolomics_processed_datasets[[nn1]] -->
<!--   x_log = d$normalized_data[[1]] -->
<!--   x_unlog = 2^x_log -->
<!--   x_meta = d$covs -->

<!--   # take the covariates, ignore distances -->
<!--   curr_covs = data.frame( -->
<!--       sex=x_meta$animal.registration.sex, -->
<!--       sample_order = x_meta$sample_order -->
<!--   ) -->

<!--   # get the lm objects -->
<!--   curr_models = list() -->
<!--   for(tp in unique(x_meta$animal.key.timepoint)){ -->
<!--       res_log = apply( -->
<!--         x_log,1, -->
<!--         pass1a_simple_differential_abundance, -->
<!--         tps = x_meta$animal.key.timepoint,tp=tp, -->
<!--         is_control = x_meta$animal.key.is_control, -->
<!--         covs = curr_covs,return_model=T -->
<!--       ) -->
<!--       res_unlog = apply( -->
<!--         x_unlog,1, -->
<!--         pass1a_simple_differential_abundance, -->
<!--         tps = x_meta$animal.key.timepoint,tp=tp, -->
<!--         is_control = x_meta$animal.key.is_control, -->
<!--         covs = curr_covs,return_model=T -->
<!--       ) -->
<!--       is_norm = cbind( -->
<!--         sapply(res_log,function(x)is_normal_test(residuals(x))), -->
<!--         sapply(res_unlog,function(x)is_normal_test(residuals(x))) -->
<!--       ) -->
<!--       colnames(is_norm) = c("log","not log") -->
<!--       curr_models[[as.character(tp)]] = is_norm -->
<!--   } -->
<!--   residual_analysis_results[[nn1]] = curr_models -->
<!-- } -->

<!--   # Is there a significant difference between the two options? -->
<!-- log_vs_unlog_summ_mat = sapply(residual_analysis_results, -->
<!--     function(x)sapply(x, -->
<!--         function(y) -->
<!--           wilcox.test(y[,1],y[,2],paired = T,alternative = "g")$p.value)) -->

<!-- # Count the number of non-normal metabolites -->
<!-- num_nonnormal_log = sapply(residual_analysis_results, -->
<!--     function(x)sapply(x, -->
<!--         function(y)sum(y[,1]<0.05))) -->
<!-- num_nonnormal_log =  -->
<!--   num_nonnormal_log[,order(colnames(num_nonnormal_log))] -->
<!-- num_nonnormal_unlog = sapply(residual_analysis_results, -->
<!--     function(x)sapply(x, -->
<!--         function(y)sum(y[,2]<0.05))) -->
<!-- num_nonnormal_unlog =  -->
<!--   num_nonnormal_unlog[,order(colnames(num_nonnormal_unlog))] -->

<!-- library(corrplot) -->
<!-- par(mar = c(5,5,5,10)) -->
<!-- normdiffs = t(num_nonnormal_log)- t(num_nonnormal_unlog) -->
<!-- corrplot(normdiffs,is.corr = F,tl.cex = 0.7) -->
<!-- ``` -->









